---
title: "3. Pathway analysis"
output: html_document
date: "2024-11-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

These scripts analyse the output "linked genes" and "persistent linked genes" from the tumor airways comparisons combined script using pathway analysis.

Performing GSEA of the datasets:

### Set up the necessary packages
```{r message=FALSE, warning=FALSE, results='hide'}

# Load the required packages
library(clusterProfiler, verbose = FALSE)
library(enrichplot, verbose = FALSE)
library(ggplot2, verbose = FALSE)
organism = "org.Hs.eg.db" # Human annotation database download
BiocManager::install(organism, character.only = TRUE, verbose = FALSE) 
library(organism, character.only = TRUE, verbose = FALSE)

```

### Load and prepare the lists of interest, here "linked genes" and "linked persistent genes" (subject to change)

2024/11/13: I am electing to rank genes by the log2FC values from the airway dataset (A1) for pathway analysis. I reason that we are more interested in the dysregulated pathways in the airways than the tumors even if we are looking at the genes linked to all 3 modalities.
```{r}

#A1_TE_TM_linked_genes <- read.table("../2_Outputs/6_Linked_genes/A1_TE_TM_linked_genes_0.01_0_0.01_1_0.01_0.5_20241211.txt", header = TRUE)

#A1_TE_TM_linked_genes <- read.table("../2_Outputs/6_Linked_genes/A1_TE_TM_linked_genes_0.05_0.2_0.05_0.8_0.05_0.5_20241211.txt", header = TRUE)

A1_TE_TM_linked_genes <- read.table("../2_Outputs/6_Linked_genes/A1_TE_TM_linked_genes_0.05_0.1_0.001_0.6_0.05_0.1_20241212.txt", header = TRUE)

# The log2FC of the airway dataset is what I am choosing for pathway analysis
linked_gene_list_A1_sorted <- A1_TE_TM_linked_genes$log2FC_A1

# name the vector
names(linked_gene_list_A1_sorted) <- A1_TE_TM_linked_genes$Gene

# sort the list in decreasing order (required for clusterProfiler)
linked_gene_list_A1_sorted = sort(linked_gene_list_A1_sorted, decreasing = TRUE)

```


### Perform gene set enrichment analysis with GO
```{r message=FALSE, warning=FALSE}
# From the default on the website, I changed pAdjustMethod to "fdr", keyType to "SYMBOL". And there was a warning message that I should remove the nPerm parameter so I did that

## overrepresentation analysis, simpliofy() function to remove redundant

linked_gene_list_A1_sorted_gse <- clusterProfiler::gseGO(geneList=linked_gene_list_A1_sorted, 
             ont ="ALL", 
             keyType = "SYMBOL", 
             # nPerm = 10000, 
             minGSSize = 3, 
             maxGSSize = 800, 
             pvalueCutoff = 0.05, 
             verbose = TRUE, 
             OrgDb = organism, 
             pAdjustMethod = "fdr")

# To view the result
linked_gene_list_A1_sorted_gse@result
```

### Make a GO dotplot of the top 10 categories of enrichment
```{r}

library(ggplot2)
require(DOSE)

linked_gene_list_A1_sorted_dotplot <- enrichplot::dotplot(linked_gene_list_A1_sorted_gse, showCategory=10, split=".sign") + facet_grid(.~.sign)

linked_gene_list_A1_sorted_dotplot


```

### Make a GO enrichment map of top 10 categories
```{r, eval = FALSE}

# Calculate term similarity matrix
term_sim_matrix <- enrichplot::pairwise_termsim(linked_gene_list_A1_sorted_gse)

# Visualize enrichment results with the term similarity matrix
linked_gene_list_A1_sorted_emapplot <- enrichplot::emapplot(term_sim_matrix, showCategory = 9)

linked_gene_list_A1_sorted_emapplot

# Save the plot
#ggsave("../2_Output/RMA_output/plots/GSEA_plots/linked_gene_list_A1_sorted/linked_gene_list_A1_sorted_GO_emap.png", linked_gene_list_A1_sorted_emapplot, width = 10, height = 10)
```

### Make a GO category netplot
```{r, eval = FALSE}
# categorySize can be either 'pvalue' or 'geneNum'
linked_gene_list_A1_sorted_GO_netplot <- enrichplot::cnetplot(linked_gene_list_A1_sorted_gse, categorySize="pvalue", foldChange=linked_gene_list_A1_sorted, showCategory = 3)

linked_gene_list_A1_sorted_GO_netplot

# Save the plot
#ggsave("../2_Output/RMA_output/plots/GSEA_plots/linked_gene_list_A1_sorted/linked_gene_list_A1_sorted_GO_cnet.png", linked_gene_list_A1_sorted_GO_netplot, width = 15, height = 10)

```

## Make a ridgeplot
```{r, fig.width = 20, fig.height = 15, eval = FALSE}
## Make a ridgeplot
linked_gene_list_A1_sorted_GO_ridgeplot <- enrichplot::ridgeplot(linked_gene_list_A1_sorted_gse) + labs(x = "enrichment distribution")

linked_gene_list_A1_sorted_GO_ridgeplot

# Save the plot
#ggsave("../2_Output/RMA_output/plots/GSEA_plots/linked_gene_list_A1_sorted/linked_gene_list_A1_sorted_GO_ridgeplot.png", linked_gene_list_A1_sorted_GO_ridgeplot, width = 20, height = 15)

```

## Make GSEA plots
```{r}
# Use the `Gene Set` param for the index in the title, and as the value for geneSetId

enrichplot::gseaplot(linked_gene_list_A1_sorted_gse, by = "all", title = linked_gene_list_A1_sorted_gse$Description[1], geneSetID = 1)

enrichplot::gseaplot(linked_gene_list_A1_sorted_gse, by = "all", title = linked_gene_list_A1_sorted_gse$Description[2], geneSetID = 2)

enrichplot::gseaplot(linked_gene_list_A1_sorted_gse, by = "all", title = linked_gene_list_A1_sorted_gse$Description[3], geneSetID = 3)
```

Interpreting GSEA plots: https://www.youtube.com/watch?v=Yi4d7JIlAsM
Alright that's it for GO, now I will try KEGG

### Prepare input for KEGG
```{r}
# Convert gene IDs for gseKEGG function
# We will lose some genes here because not all IDs will be converted
ids<-clusterProfiler::bitr(names(linked_gene_list_A1_sorted), fromType = "SYMBOL", toType = "ENTREZID", OrgDb=organism)
 # remove duplicate IDS (here I use "ENSEMBL", but it should be whatever was selected as keyType)
dedup_ids = ids[!duplicated(ids[c("SYMBOL")]),]

# Create a new vector which has only the genes which were successfully mapped using the bitr function above
linked_gene_list_A1_sorted_kegg = linked_gene_list_A1_sorted[names(linked_gene_list_A1_sorted) %in% dedup_ids$SYMBOL]

linked_gene_list_A1_sorted_kegg <- linked_gene_list_A1_sorted_kegg 

# Name vector with ENTREZ ids
names(linked_gene_list_A1_sorted_kegg) <- dedup_ids$ENTREZID

# sort the list in decreasing order (required for clusterProfiler)
linked_gene_list_A1_sorted_kegg = sort(linked_gene_list_A1_sorted_kegg, decreasing = TRUE)
```

### Run KEGG pathway analysis:
```{r}
kegg_organism = "hsa"
linked_gene_list_A1_sorted_kegg_gse <- clusterProfiler::gseKEGG(geneList = linked_gene_list_A1_sorted_kegg,
               organism     = kegg_organism,
              # nPerm        = 10000,
               minGSSize    = 3,
               maxGSSize    = 800,
               pvalueCutoff = 0.05,
               pAdjustMethod = "fdr",
               keyType       = "ncbi-geneid")

linked_gene_list_A1_sorted_kegg_gse@result
```

## Make KEGG dotplot

```{r}
linked_gene_list_A1_sorted_kegg_dotplot <- enrichplot::dotplot(linked_gene_list_A1_sorted_kegg_gse, showCategory = 10, title = "Enriched Pathways" , split=".sign") + facet_grid(.~.sign)

linked_gene_list_A1_sorted_kegg_dotplot

#ggsave("../2_Output/RMA_output/plots/GSEA_plots/linked_gene_list_A1_sorted/linked_gene_list_A1_sorted_kegg_dotplot.png", linked_gene_list_A1_sorted_kegg_dotplot, width = 7, height = 12)
```


# Make a kegg enrichment map of top 10 categories
```{r, fig.width = 10, fig.height = 10, eval = FALSE}

# Calculate term similarity matrix
term_sim_matrix <- enrichplot::pairwise_termsim(linked_gene_list_A1_sorted_kegg_gse)

# Visualize enrichment results with the term similarity matrix
linked_gene_list_A1_sorted_kegg_emapplot <- enrichplot::emapplot(term_sim_matrix, showCategory = 10)
linked_gene_list_A1_sorted_kegg_emapplot

# Save the plot
#ggsave("../2_Output/RMA_output/plots/GSEA_plots/linked_gene_list_A1_sorted/linked_gene_list_A1_sorted_kegg_emap.png", linked_gene_list_A1_sorted_kegg_emapplot, width = 10, height = 10)
## Make a kegg category netplot
```



```{r, fig.width = 15, fig.height = 10, eval = FALSE}
# categorySize can be either 'pvalue' or 'geneNum'
linked_gene_list_A1_sorted_kegg_netplot <- enrichplot::cnetplot(linked_gene_list_A1_sorted_kegg_gse, categorySize="pvalue", foldChange=linked_gene_list_A1_sorted, showCategory = 3)

linked_gene_list_A1_sorted_kegg_netplot

# Save the plot
#ggsave("../2_Output/RMA_output/plots/GSEA_plots/linked_gene_list_A1_sorted/linked_gene_list_A1_sorted_kegg_cnet.png", linked_gene_list_A1_sorted_kegg_netplot, width = 15, height = 10)

```

### Make a kegg ridgeplot
```{r, fig.width = 20, fig.height = 15, eval = FALSE}

## Make a ridgeplot
linked_gene_list_A1_sorted_kegg_ridgeplot<- enrichplot::ridgeplot(linked_gene_list_A1_sorted_kegg_gse) + labs(x = "enrichment distribution")

linked_gene_list_A1_sorted_kegg_ridgeplot

# Save the plot
#ggsave("../2_Output/RMA_output/plots/GSEA_plots/linked_gene_list_A1_sorted/linked_gene_list_A1_sorted_kegg_ridgeplot.png", linked_gene_list_A1_sorted_kegg_ridgeplot, width = 20, height = 15)

```

