---
title: "5_Airway_CS_NS_meta_analysis"
output: html_document
date: "2024-11-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

These scripts analyse the output "linked genes" and "persistent linked genes" from the tumor airways comparisons combined script, and use a meta-analysis approach to give supporting evidence that they are differentially expressed in airways of smokers.

# Making a combined table to rank genes of interest and compare studies

### Read in the linked genes and persistent linked genes
```{r}

# Read in the linked genes
linked_genes <- read.table('../2_Outputs/A1_TE_TM_linked_genes_nocutoffs_20241112.txt', header = TRUE)

# Read in the persistent linked genes
persistent_linked_genes <- read.table('../2_Outputs/A1_TE_TM_A2_persistent_linked_genes_nocutoffs_20241112.txt', header = TRUE)

```

## Read in airway DEGs from all datasets
```{r}
## Read in the tables

GSE4302_CS_NS_GEO2R_limma_sig <- read.table("../2_Outputs/1_Airway_DEGs/GSE4302_CS_NS_GEO2R_limma_sig_20241114.txt", sep = '\t')

GSE7895_CS_NS_GEO2R_limma_sig <- read.table("../2_Outputs/1_Airway_DEGs/GSE7895_CS_NS_GEO2R_limma_sig_20241114.txt", sep = '\t')

GSE14633_CS_NS_GEO2R_limma_sig <- read.table("../2_Outputs/1_Airway_DEGs/GSE14633_CS_NS_GEO2R_limma_sig_20241114.txt", sep = '\t')
colnames(GSE14633_CS_NS_GEO2R_limma_sig)[2] <- "Gene.symbol" # Rename column for binding

GSE19027_CS_NS_GEO2R_limma_sig <- read.table("../2_Outputs/1_Airway_DEGs/GSE19027_CS_NS_GEO2R_limma_sig_20241114.txt", sep = '\t')
# For some reason I have to set the column names
colnames(GSE19027_CS_NS_GEO2R_limma_sig) <- GSE19027_CS_NS_GEO2R_limma_sig[1,]
GSE19027_CS_NS_GEO2R_limma_sig <- GSE19027_CS_NS_GEO2R_limma_sig[-1,]

GSE20257_CS_NS_GEO2R_limma_sig <- read.table("../2_Outputs/1_Airway_DEGs/GSE20257_CS_NS_GEO2R_limma_sig_20241114.txt", sep = '\t')
# For some reason I have to set the column names
colnames(GSE20257_CS_NS_GEO2R_limma_sig) <- GSE20257_CS_NS_GEO2R_limma_sig[1,]
GSE20257_CS_NS_GEO2R_limma_sig <- GSE20257_CS_NS_GEO2R_limma_sig[-1,]

GSE63127_CS_NS_GEO2R_limma_sig <- read.table("../2_Outputs/1_Airway_DEGs/GSE63127_CS_NS_GEO2R_limma_sig_20241115.txt", sep = '\t')
# For some reason I have to set the column names
colnames(GSE63127_CS_NS_GEO2R_limma_sig) <- GSE63127_CS_NS_GEO2R_limma_sig[1,]
GSE63127_CS_NS_GEO2R_limma_sig <- GSE63127_CS_NS_GEO2R_limma_sig[-1,]

```

## Make the combined table
```{r}
# Combine DEG_sig lists together, with study ID identifier column
library(dplyr)

CS_NS_gene_rank_input <- as.data.frame(rbind(
  cbind("GSE19027", GSE19027_CS_NS_GEO2R_limma_sig$Gene.symbol, GSE19027_CS_NS_GEO2R_limma_sig$logFC, GSE19027_CS_NS_GEO2R_limma_sig$adj.P.Val ), 
  cbind("GSE14633", GSE14633_CS_NS_GEO2R_limma_sig$Gene.symbol, GSE14633_CS_NS_GEO2R_limma_sig$logFC, GSE14633_CS_NS_GEO2R_limma_sig$adj.P.Val), 
  cbind("GSE20257", GSE20257_CS_NS_GEO2R_limma_sig$Gene.symbol, GSE20257_CS_NS_GEO2R_limma_sig$logFC, GSE20257_CS_NS_GEO2R_limma_sig$adj.P.Val), 
  cbind("GSE4302", GSE4302_CS_NS_GEO2R_limma_sig$Gene.symbol, GSE4302_CS_NS_GEO2R_limma_sig$logFC, GSE4302_CS_NS_GEO2R_limma_sig$adj.P.Val), 
  cbind("GSE63127",GSE63127_CS_NS_GEO2R_limma_sig$Gene.symbol, GSE63127_CS_NS_GEO2R_limma_sig$logFC, GSE63127_CS_NS_GEO2R_limma_sig$adj.P.Val), 
  cbind("GSE7895", GSE7895_CS_NS_GEO2R_limma_sig$Gene.symbol, GSE7895_CS_NS_GEO2R_limma_sig$logFC, GSE7895_CS_NS_GEO2R_limma_sig$adj.P.Val)))

colnames(CS_NS_gene_rank_input) <- c("Study_ID", "Gene.symbol", "log2FC", "adj.P.Val")
head(CS_NS_gene_rank_input)

```

## Creating the ranked list
```{r}

# Convert the log2FC and adj.P.Val to numeric values
CS_NS_gene_rank_input$log2FC <- as.numeric(CS_NS_gene_rank_input$log2FC)
CS_NS_gene_rank_input$adj.P.Val <- as.numeric(CS_NS_gene_rank_input$adj.P.Val)

# Easier variable name
df <- CS_NS_gene_rank_input

# Step 1: Group by gene name
grouped_df <- df %>%
  group_by(Gene.symbol)

# Step 2: Create a summary table with the required statistics
summary_table <- grouped_df %>%
  summarise(
    num_studies = n_distinct(Study_ID),
    studies_list = paste(unique(Study_ID), collapse = ", "),
    avg_log2FC = mean(log2FC, na.rm = TRUE),
    avg_abs_log2FC = abs(mean(log2FC, na.rm = TRUE)),
    sd_log2FC = sd(log2FC),
    avg_adj.P.Val = mean(adj.P.Val, na.rm = TRUE),
    sd_adj.P.Val = sd(adj.P.Val)
  )

# Normalize values to a common scale (0 to 1)
summary_table$normalized_num_studies <- (summary_table$num_studies - min(summary_table$num_studies)) / (max(summary_table$num_studies) - min(summary_table$num_studies))
summary_table$normalized_avg_abs_log2FC <- (summary_table$avg_abs_log2FC - min(summary_table$avg_abs_log2FC)) / (max(summary_table$avg_abs_log2FC) - min(summary_table$avg_abs_log2FC))

# Assign weights for both factors (modify as needed)
num_studies_weight <- 0.5
abs_log2FC_weight <- 0.5

# Calculate the ranking score
summary_table$ranking_score <- num_studies_weight * summary_table$normalized_num_studies +
                               abs_log2FC_weight * summary_table$normalized_avg_abs_log2FC

# Rank the genes based on the calculated ranking score
CS_NS_ranked_genes <- summary_table %>%
  arrange(desc(ranking_score))

# Add a column with the rank order
CS_NS_ranked_genes["rank_order"] <- rownames(CS_NS_ranked_genes)
#head(CS_NS_ranked_genes, n = 20)

# Save the output
write.table(CS_NS_ranked_genes, "../2_Outputs/1_Airway_DEGs/CS_NS_GEO2R_limma_ranked_genes.txt", sep = '\t', row.names = FALSE, col.names = TRUE)

```

## Histograms showing numbers of genes occurring in multiple studies (overall genes and genes of interest)

```{r}

# Histogram of number of studies every gene appears in
hist(CS_NS_ranked_genes$num_studies, 
     xlim = c(0.5, 6.5),  # Adjust x-axis limits slightly
     ylim = c(0,6000),
     breaks = seq(0.5, 6.5, by = 1),  # Shift breaks by 0.5
     main = "Number of CS/NS airway datasets all genes are significant in",
     xlab = "Number of datasets",
     ylab = "Number of genes")

# Filter the ranked genes to the "linked" genes
CS_NS_ranked_genes_linked <- CS_NS_ranked_genes %>%
  filter(Gene.symbol %in% linked_genes$Gene)
# Histogram of number of studies every "linked" gene appears in
hist(CS_NS_ranked_genes_linked$num_studies, 
     xlim = c(0.5, 6.5),  # Adjust x-axis limits slightly
     #ylim = c(0,6000),
     breaks = seq(0.5, 6.5, by = 1),  # Shift breaks by 0.5
     main = "Number of CS/NS airway datasets the 'linked' genes are significant in",
     xlab = "Number of datasets",
     ylab = "Number of genes")

# Filter the ranked genes to the persistent linked genes
CS_NS_ranked_genes_persistent_linked <- CS_NS_ranked_genes %>%
  filter(Gene.symbol %in% persistent_linked_genes$Gene)
# Histogram of number of studies every "persistent linked" gene appears in
hist(CS_NS_ranked_genes_persistent_linked$num_studies, 
     xlim = c(0.5, 6.5),  # Adjust x-axis limits slightly
     #ylim = c(0,6000),
     breaks = seq(0.5, 6.5, by = 1),  # Shift breaks by 0.5
     main = "Number of CS/NS airway datasets the 'persistent linked' genes are significant in",
     xlab = "Number of datasets",
     ylab = "Number of genes")

```


# Calculating standard mean difference (SMD)

### getMeanSD: Function to return a table of mean and SD values for each gene in a matrix. 
```{r}
getMeanSD <- function(GSE_readcount_input, current_smoker_indexes, never_smoker_indexes){

	# Make dataframe to contain the info
	GSE_CS_NS_mean_SD <- data.frame(
	  ID = rownames(GSE_readcount_input),
	  Current_smoker_count = NA,
	  Current_smoker_mean = NA,
	  Current_smoker_SD = NA,
	  Never_smoker_count = NA,
	  Never_smoker_mean = NA,
	  Never_smoker_SD = NA
	)
	
	# Loop through each gene and calculate mean and standard deviation
	for (i in 1:nrow(GSE_readcount_input)) { #Looping through rows
	  Current_row <- GSE_readcount_input[i,] #Extract the row
	  current_smoker_values <- GSE_readcount_input[i, current_smoker_indexes]
	  never_smoker_values   <- GSE_readcount_input[i, never_smoker_indexes]
	  
	  # Calculate mean and standard deviation for Current_smoker
	  current_mean <- mean(as.numeric(current_smoker_values))
	  current_sd   <- sd(as.numeric(current_smoker_values))
	  
	  # Calculate mean and standard deviation for Never_smoker
	  never_mean <- mean(as.numeric(never_smoker_values))
	  never_sd   <- sd(as.numeric(never_smoker_values))
	  
	  # Add results to the new dataframe
	  GSE_CS_NS_mean_SD[i, 2:4] <- c(length(current_smoker_values), current_mean, current_sd)
	  GSE_CS_NS_mean_SD[i, 5:7] <- c(length(never_smoker_values), never_mean, never_sd)
	}
	return(GSE_CS_NS_mean_SD)
}

# Save the function
#dump("getMeanSD", file = "./Functions/getMeanSD.R")
```

### Read in the normalized read count information
```{r}
GSE4302_expr_matrix  <- read.table("../2_Outputs/2_Airway_expression/GSE4302_expr_matrix_20241115.txt", header = T)
GSE4302_sstat_data   <- read.table("../2_Outputs/2_Airway_expression/GSE4302_sstat_data.txt", header = T)
GSE4302_gene_symbol  <- read.table("../2_Outputs/2_Airway_expression/GSE4302_gene_symbol.txt", header = T)

GSE7895_expr_matrix  <- read.table("../2_Outputs/2_Airway_expression/GSE7895_expr_matrix_20241115.txt", header = T)
GSE7895_sstat_data   <- read.table("../2_Outputs/2_Airway_expression/GSE7895_sstat_data.txt", header = T)
GSE7895_gene_symbol  <- read.table("../2_Outputs/2_Airway_expression/GSE7895_gene_symbol.txt", header = T)

GSE14633_expr_matrix <- read.table("../2_Outputs/2_Airway_expression/GSE14633_expr_matrix_20241115.txt", header = T)
GSE14633_sstat_data  <- read.table("../2_Outputs/2_Airway_expression/GSE14633_sstat_data.txt", header = T)
GSE14633_gene_symbol <- read.table("../2_Outputs/2_Airway_expression/GSE14633_gene_symbol.txt", header = T)

GSE19027_expr_matrix <- read.table("../2_Outputs/2_Airway_expression/GSE19027_expr_matrix_20241115.txt", header = T)
GSE19027_sstat_data  <- read.table("../2_Outputs/2_Airway_expression/GSE19027_sstat_data.txt", header = T)
GSE19027_gene_symbol <- read.table("../2_Outputs/2_Airway_expression/GSE19027_gene_symbol.txt", header = T)

GSE20257_expr_matrix <- read.table("../2_Outputs/2_Airway_expression/GSE20257_expr_matrix_20241115.txt", header = T)
GSE20257_sstat_data  <- read.table("../2_Outputs/2_Airway_expression/GSE20257_sstat_data.txt", header = T)
GSE20257_gene_symbol <- read.table("../2_Outputs/2_Airway_expression/GSE20257_gene_symbol.txt", header = T)

GSE63127_expr_matrix <- read.table("../2_Outputs/2_Airway_expression/GSE63127_expr_matrix_20241118.txt", header = T)
GSE63127_sstat_data  <- read.table("../2_Outputs/2_Airway_expression/GSE63127_sstat_data.txt", header = T)
GSE63127_gene_symbol  <- read.table("../2_Outputs/2_Airway_expression/GSE63127_gene_symbol.txt", header = T)

```

### Apply getMeanSD function to CS vs NS comparisons, and add GEO_ID identifier column
```{r}

## GSE4302
GSE4302_current_smoker_indexes <- which(GSE4302_sstat_data[1]=="current_smoker")
GSE4302_never_smoker_indexes   <- which(GSE4302_sstat_data[1]=="healthy_control")
GSE4302_CS_NS_mean_SD <- getMeanSD(GSE4302_expr_matrix, GSE4302_current_smoker_indexes, GSE4302_never_smoker_indexes)
GSE4302_CS_NS_mean_SD <- cbind(GEO_ID = "GSE4302", GSE4302_CS_NS_mean_SD)
GSE4302_CS_NS_mean_SD$Gene <- GSE4302_gene_symbol$x # Add gene symbol column

# GSE7895
GSE7895_current_smoker_indexes <- which(GSE7895_sstat_data[1]=="current_smoker")
GSE7895_never_smoker_indexes   <- which(GSE7895_sstat_data[1]=="never_smoker")
GSE7895_CS_NS_mean_SD <- getMeanSD(GSE7895_expr_matrix, GSE7895_current_smoker_indexes, GSE7895_never_smoker_indexes)
GSE7895_CS_NS_mean_SD <- cbind(GEO_ID = "GSE7895", GSE7895_CS_NS_mean_SD)
GSE7895_CS_NS_mean_SD$Gene <- GSE7895_gene_symbol$x

# GSE19027
GSE19027_current_smoker_indexes <- which(GSE19027_sstat_data[1]=="current_smoker")
GSE19027_never_smoker_indexes   <- which(GSE19027_sstat_data[1]=="never_smoker")
GSE19027_CS_NS_mean_SD <- getMeanSD(GSE19027_expr_matrix, GSE19027_current_smoker_indexes, GSE19027_never_smoker_indexes)
GSE19027_CS_NS_mean_SD <- cbind(GEO_ID = "GSE19027", GSE19027_CS_NS_mean_SD)
GSE19027_CS_NS_mean_SD$Gene <- GSE19027_gene_symbol$x

# GSE20257
GSE20257_current_smoker_indexes <- which(GSE20257_sstat_data[1]=="smoker")
GSE20257_never_smoker_indexes   <- which(GSE20257_sstat_data[1]=="non.smoker")
GSE20257_CS_NS_mean_SD <- getMeanSD(GSE20257_expr_matrix, GSE20257_current_smoker_indexes, GSE20257_never_smoker_indexes)
GSE20257_CS_NS_mean_SD <- cbind(GEO_ID = "GSE20257", GSE20257_CS_NS_mean_SD)
GSE20257_CS_NS_mean_SD$Gene <- GSE20257_gene_symbol$x

# GSE63127
GSE63127_current_smoker_indexes <- which(GSE63127_sstat_data[1]=="smoker")
GSE63127_never_smoker_indexes   <- which(GSE63127_sstat_data[1]=="non.smoker")
GSE63127_CS_NS_mean_SD <- getMeanSD(GSE63127_expr_matrix, GSE63127_current_smoker_indexes, GSE63127_never_smoker_indexes)
GSE63127_CS_NS_mean_SD <- cbind(GEO_ID = "GSE63127", GSE63127_CS_NS_mean_SD)
GSE63127_CS_NS_mean_SD$Gene <- GSE63127_gene_symbol$x

```

### Apply getMeanSD function to the persistent linked gene list to generate a table of SMD values
```{r}
library(meta)

## Use the persistent linked genes list as the input genes list
gene_list <- persistent_linked_genes

## Put the dataframes in a list
dataframe_list <- list(GSE4302_CS_NS_mean_SD, GSE7895_CS_NS_mean_SD, GSE19027_CS_NS_mean_SD, GSE20257_CS_NS_mean_SD, GSE63127_CS_NS_mean_SD)


# Make a dataframe as a template that will hold the values for each gene as input to SMD calculation
CS_NS_SMD_input_template <- data.frame(
	  GEO_ID = character(0),
    Gene = character(0),
	  Current_smoker_count = numeric(0),
	  Current_smoker_mean = numeric(0),
	  Current_smoker_SD = numeric(0),
	  Never_smoker_count = numeric(0),
	  Never_smoker_mean = numeric(0),
	  Never_smoker_SD = numeric(0)
	)
# Make a dataframe that will have all the SMD output values
persistent_linked_CS_NS_SMD_summary <- data.frame(
  GEO_ID = character(0),
  gene = character(0),
  ID = character(0),
  n_obs = numeric(0),
  SMD = numeric(0),
  CI_95_lower = numeric(0),
  CI_95_upper = numeric(0),
  z_statistic = numeric(0),
  p_value = numeric(0),
  method = character(0)
)

# Loop through the list and extract individual genes
for (i in 1:nrow(gene_list)){ # For each gene on the gene list
  
  df_SMD_input <- CS_NS_SMD_input_template # Make a new dataframe for data to input into SMD
  
  for (j in 1:length(dataframe_list)){ # For each dataframe in the list of dataframes
    df <- dataframe_list[[j]] # Extract dataframe j
    # Append the row with the gene of interest to the SMD input dataframe
    df_SMD_input <- rbind(df_SMD_input, df[df$Gene==gene_list[i,]$Gene,])
  }
  
  for (k in 1:length(df_SMD_input[,1])){ # For all the GSE values in the input
    # Input the resulting df into the SMD calculation function
    df_SMD_values <- metacont(Current_smoker_count, Current_smoker_mean, Current_smoker_SD, Never_smoker_count, Never_smoker_mean, Never_smoker_SD, sm="SMD", data = df_SMD_input, subset=k)
  
  
  # Make a summary of the values
    df_SMD_summary <- as.data.frame(t(c(GEO_ID = df_SMD_input[k,1], gene = gene_list$Gene[i], ID = df_SMD_input$ID[k], n_obs = sum(df_SMD_values$n.e, df_SMD_values$n.c), SMD = df_SMD_values$TE, CI_95_lower = df_SMD_values$lower, CI_95_upper =  df_SMD_values$upper, z_stat = df_SMD_values$statistic, p_value = df_SMD_values$pval, method = df_SMD_values$method.smd)))
  
  # Append results from this to the output dataframe
    persistent_linked_CS_NS_SMD_summary <- rbind(persistent_linked_CS_NS_SMD_summary, df_SMD_summary)
  }
}

head(persistent_linked_CS_NS_SMD_summary)
```

## Filter the SMD table to the probesets of interest (those that were selected in the DEGs calculations due to having lowest FDR)
```{r}

## Grab all the probe IDs of interest

dataframe_DEGs_list <- list(GSE4302 = GSE4302_CS_NS_GEO2R_limma_sig, GSE63127 = GSE63127_CS_NS_GEO2R_limma_sig, GSE14633 = GSE14633_CS_NS_GEO2R_limma_sig, GSE19027 = GSE19027_CS_NS_GEO2R_limma_sig, GSE20257 = GSE20257_CS_NS_GEO2R_limma_sig, GSE7895 = GSE7895_CS_NS_GEO2R_limma_sig)

GSE_probeset_gene <- data.frame(GSE = character(0), probeset = character(0), gene = character(0))


## I worked through the logic for this in my head and it is difficult to explain in words
for (i in 1:length(dataframe_DEGs_list)) { # Iterate through all the dataframes of DEGs
  df <- dataframe_DEGs_list[[i]]
  for (j in 1:nrow(gene_list)){
    df_gene_j <- df[df$Gene==gene_list$Gene[j],] # Pull out the rows with gene names in the persistent linked genes list
    df_name <- names(dataframe_DEGs_list[i])
    if (length(df_gene_j$ID)>0) { # If that row existed in this dataframe
      probeset_ID_gene_j <- df_gene_j$ID
      gene_j <- df_gene_j$Gene.symbol
      GSE_probeset_pair_gene_j <- data.frame(df_name, probeset_ID_gene_j, gene_j)
      GSE_probeset_gene <- rbind(GSE_probeset_pairs, GSE_probeset_pair_gene_j) # Store the GEO ID, probeset ID, and gene name
    }  
  }
}

colnames(GSE_probeset_gene) <- c("GEO_ID", "ID", "gene")

# Now see if this works for filtering the main list
persistent_linked_CS_NS_SMD_summary_ID_filter <- persistent_linked_CS_NS_SMD_summary %>%
  semi_join(GSE_probeset_gene, by = c("GEO_ID", "gene", "ID"))

```

## Cumulative SMDs for each gene (overall summary for a gene, not separated by study)

```{r}

library(meta)

## Use the persistent linked genes list as the input genes list
gene_list <- persistent_linked_genes

## Put the dataframes in a list
dataframe_list <- list(GSE4302 = GSE4302_CS_NS_mean_SD, GSE7895 = GSE7895_CS_NS_mean_SD, GSE19027 = GSE19027_CS_NS_mean_SD, GSE20257 = GSE20257_CS_NS_mean_SD, GSE63127 = GSE63127_CS_NS_mean_SD)

# Make a dataframe as a template that will hold the values for each gene as input to SMD calculation
CS_NS_SMD_input_template <- data.frame(
	  GEO_ID = character(0),
    Gene = character(0),
	  Current_smoker_count = numeric(0),
	  Current_smoker_mean = numeric(0),
	  Current_smoker_SD = numeric(0),
	  Never_smoker_count = numeric(0),
	  Never_smoker_mean = numeric(0),
	  Never_smoker_SD = numeric(0)
	)
# Make a dataframe that will have all the SMD output values
persistent_linked_CS_NS_SMD_summary_overall <- data.frame(
  GEO_ID = character(0),
  gene = character(0),
  n_obs = numeric(0),
  SMD = numeric(0),
  CI_95_lower = numeric(0),
  CI_95_upper = numeric(0),
  z_statistic = numeric(0),
  p_value = numeric(0),
  method = character(0)
)

# Making this to help filter to the probesets of interest
GSE_probeset_gene_for_join <- GSE_probeset_gene
names(GSE_probeset_gene_for_join)[3] <- "Gene" 

# Loop through the list and extract individual genes
for (i in 1:nrow(gene_list)){ # For each gene on the gene list
  
  df_SMD_input <- CS_NS_SMD_input_template # Make a new dataframe for data to input into SMD
  
  for (j in 1:length(dataframe_list)){ # For each dataframe in the list of dataframes
    df <- dataframe_list[[j]] # Extract dataframe j
    df_gene_j <- df[df$Gene==gene_list[i,]$Gene,]
    df_gene_j <- df_gene_j %>%
      semi_join(GSE_probeset_gene_for_join, by = c("GEO_ID", "Gene", "ID")) # Filtering to probesets of interest from DEGs
    # Append the row with the gene of interest to the SMD input dataframe
    df_SMD_input <- rbind(df_SMD_input, df_gene_j)
  }
  
  # Input the resulting df into the SMD calculation function
  df_SMD_values <- metacont(Current_smoker_count, Current_smoker_mean, Current_smoker_SD, Never_smoker_count, Never_smoker_mean, Never_smoker_SD, sm="SMD", data = df_SMD_input)
  
  
  # Make a summary of the values
      df_SMD_summary_overall <- as.data.frame(t(c(gene = gene_list$Gene[i], n_studies = df_SMD_values$k, n_obs = sum(df_SMD_values$n.e.pooled, df_SMD_values$n.c.pooled), SMD_common= df_SMD_values$TE.common, CI_95_lower_common = df_SMD_values$lower.common, CI_95_upper_common =  df_SMD_values$upper.common, z_stat_common = df_SMD_values$statistic.common, p_value_common = df_SMD_values$pval.common, SMD_random= df_SMD_values$TE.random, CI_95_lower_random = df_SMD_values$lower.random, CI_95_upper_random =  df_SMD_values$upper.random, z_stat_random = df_SMD_values$statistic.random, p_value_random = df_SMD_values$pval.random, method = df_SMD_values$method.smd, Q = df_SMD_values$Q, Q_df = df_SMD_values$df.Q,  Q_pval = df_SMD_values$pval.Q)))
  
  # Append results from this to the output dataframe
  persistent_linked_CS_NS_SMD_summary_overall <- rbind(persistent_linked_CS_NS_SMD_summary_overall, df_SMD_summary_overall)
  
}

head(persistent_linked_CS_NS_SMD_summary_overall)

```

## Forest plots

```{r, fig.width = 12, fig.height = 5}

for (i in 1:nrow(gene_list)){

  # i <- 3 # Change i to change the gene, or make a loop to check all the genes
  
  df_SMD_input <- CS_NS_SMD_input_template # Make a new dataframe for data to input into SMD
    
  for (j in 1:length(dataframe_list)){ # For each dataframe in the list of dataframes
    df <- dataframe_list[[j]] # Extract dataframe j
    df_gene_j <- df[df$Gene==gene_list[i,]$Gene,]
    df_gene_j <- df_gene_j %>%
      semi_join(GSE_probeset_gene_for_join, by = c("GEO_ID", "Gene", "ID")) # Filtering to probesets of interest from DEGs
    # Append the row with the gene of interest to the SMD input dataframe
    df_SMD_input <- rbind(df_SMD_input, df_gene_j)
  }
    
  # Input the resulting df into the SMD calculation function
  df_SMD_values <- metacont(Current_smoker_count, Current_smoker_mean, Current_smoker_SD, Never_smoker_count, Never_smoker_mean, Never_smoker_SD, sm="SMD", data = df_SMD_input, studlab = GEO_ID)
      
  # Make a forest plot of the results if the P value is less than 0.05
  if (!is.na(df_SMD_values$pval.Q) && df_SMD_values$pval.Q < 0.05){
    df_SMD_forest <- forest(df_SMD_values, smlab = paste0(gene_list$Gene[i], " SMD forest plot"))
  }
  
  # Assign a label with the name of the gene to the plot
  # file_name <- paste0(gene_list[i], "_SMD_forest")
  # assign(file_name, df_SMD_forest)

}

```

2024/11/19: So 6 of those 27 persistent linked genes have significant cumulative SMD across the airway datasets (p < 0.05)

2024/11/20: Today I could do the same process with the linked genes that are not necessarily persistent, and I could also make boxplots for genes of interest. I will start off doing boxplots of the 6 of 27 persistent linked genes that came out as having significant cumulative SMD.

## Boxplots of persistent linked genes with significant cumulative SMD across datasets

### Formatting the expression matrixes for the boxplots
```{r}

# So I want to have separate dataframes for the 5 GEO IDs, containing a column for the gene name, the expression data, and the group membership information.


# Make a list of the genes of interest for the boxplots
genes_of_interest <- c("CA12", "CCDC81", "CTNNAL1", "EFEMP1", "LMO2", "TNS1")

### GSE4302 ###
GSE4302_expr_matrix_gene <- cbind(gene = GSE4302_gene_symbol$x, GSE4302_expr_matrix)

GSE4302_gene_probe_of_interest <- GSE4302_CS_NS_GEO2R_limma_sig %>%
  filter(Gene.symbol %in% genes_of_interest) %>%
  select(Gene.symbol, ID)

GSE4302_CS_NS_expr_matrix_genes_of_interest <- GSE4302_expr_matrix_gene %>%
  filter(rownames(.) %in% GSE4302_gene_probe_of_interest$ID)


### GSE7895 ###
GSE7895_expr_matrix_gene <- cbind(gene = GSE7895_gene_symbol$x, GSE7895_expr_matrix)

GSE7895_gene_probe_of_interest <- GSE7895_CS_NS_GEO2R_limma_sig %>%
  filter(Gene.symbol %in% genes_of_interest) %>%
  select(Gene.symbol, ID)

GSE7895_CS_NS_expr_matrix_genes_of_interest <- GSE7895_expr_matrix_gene %>%
  filter(rownames(.) %in% GSE7895_gene_probe_of_interest$ID)


# GSE19027
GSE19027_expr_matrix_gene <- cbind(gene = GSE19027_gene_symbol$x, GSE19027_expr_matrix)

GSE19027_gene_probe_of_interest <- GSE19027_CS_NS_GEO2R_limma_sig %>%
  filter(Gene.symbol %in% genes_of_interest) %>%
  select(Gene.symbol, ID)

GSE19027_CS_NS_expr_matrix_genes_of_interest <- GSE19027_expr_matrix_gene %>%
  filter(rownames(.) %in% GSE19027_gene_probe_of_interest$ID)


# GSE20257
GSE20257_expr_matrix_gene <- cbind(gene = GSE20257_gene_symbol$x, GSE20257_expr_matrix)

GSE20257_gene_probe_of_interest <- GSE20257_CS_NS_GEO2R_limma_sig %>%
  filter(Gene.symbol %in% genes_of_interest) %>%
  select(Gene.symbol, ID)

GSE20257_CS_NS_expr_matrix_genes_of_interest <- GSE20257_expr_matrix_gene %>%
  filter(rownames(.) %in% GSE20257_gene_probe_of_interest$ID)


# GSE63127
GSE63127_expr_matrix_gene <- cbind(gene = GSE63127_gene_symbol$x, GSE63127_expr_matrix)

GSE63127_gene_probe_of_interest <- GSE63127_CS_NS_GEO2R_limma_sig %>%
  filter(Gene.symbol %in% genes_of_interest) %>%
  select(Gene.symbol, ID)

GSE63127_CS_NS_expr_matrix_genes_of_interest <- GSE63127_expr_matrix_gene %>%
  filter(rownames(.) %in% GSE63127_gene_probe_of_interest$ID)
```



### Iteratively creating boxplots for the genes of interest in each dataset
```{r, fig.width = 8, fig.height = 5}

library(ggplot2)
library(tidyr)

## Put the dataframes in a list
df_list_goi <- list(GSE4302 = GSE4302_CS_NS_expr_matrix_genes_of_interest, GSE7895 = GSE7895_CS_NS_expr_matrix_genes_of_interest, GSE19027 = GSE19027_CS_NS_expr_matrix_genes_of_interest, GSE20257 = GSE20257_CS_NS_expr_matrix_genes_of_interest, GSE63127 = GSE63127_CS_NS_expr_matrix_genes_of_interest)

## Put the indexes into lists
current_smoker_indexes <- list(GSE4302_csi = GSE4302_current_smoker_indexes, GSE7895_csi = GSE7895_current_smoker_indexes, GSE19027_csi = GSE19027_current_smoker_indexes, GSE20257_csi = GSE20257_current_smoker_indexes, GSE63127_csi = GSE63127_current_smoker_indexes)

never_smoker_indexes <- list(GSE4302_nsi = GSE4302_never_smoker_indexes, GSE7895_nsi = GSE7895_never_smoker_indexes, GSE19027_nsi = GSE19027_never_smoker_indexes, GSE20257_nsi = GSE20257_never_smoker_indexes, GSE63127_nsi = GSE63127_never_smoker_indexes)

# Set current gene of interest (make a loop eventually)

for (i in 1:length(genes_of_interest)){
  
  curr_gene <- genes_of_interest[i] # Loop through indexes of genes of interest
  
  for (j in 1:length(df_list_goi)){
    
    # Subset data for current smokers and never smokers
    df_curr <- as.data.frame(df_list_goi[j])
    colnames(df_curr) <- sub(".*\\.", "", colnames(df_curr))
    
    csi_curr <- unlist(current_smoker_indexes[j])
    nsi_curr <- unlist(never_smoker_indexes[j])
    
    if(curr_gene %in% df_curr$gene){
      expr_curr_gene <- df_curr[df_curr$gene==curr_gene,]
      expr_curr_gene <- expr_curr_gene[,-1]
      
      expr_current_smokers_curr_gene <- expr_curr_gene[,csi_curr]
      expr_never_smokers_curr_gene <- expr_curr_gene[,nsi_curr]
      
      # Combine into a long format dataframe
      expr_long <- data.frame(
        Gene = curr_gene,
        current_smokers = as.vector(expr_current_smokers_curr_gene),
        never_smokers   = as.vector(expr_never_smokers_curr_gene)
      ) %>%
        pivot_longer(cols = -Gene, names_to = "Group", values_to = "Expression")
      
      expr_long$Group <- sub("\\..*", "", expr_long$Group) # Remove the sample ID info
      
      # Plot using ggplot2
      boxplot<-ggplot(expr_long, aes(x = Group, y = Expression, fill = Group)) +
        geom_boxplot() +
        #facet_wrap(~ Gene, scales = "free_y") +  # One boxplot per gene
        theme_minimal() +
        scale_y_continuous(
          #limits = c(0,14), 
           minor_breaks = NULL) + 
        labs(title = paste0(names(df_list_goi[j]), " ", curr_gene, ": Current vs Never Smokers expression"),
             x = "Group", y = "Expression") +
        theme(legend.position = "none")
      print(boxplot)
    } else{
      print(paste0("Gene ", curr_gene, " not in the dataset ", names(df_list_goi[j])))
    }
  }
}
```

### Facet-wrapping boxplots for each gene of interest in all datasets
```{r, fig.width = 8, fig.height = 5}

library(ggplot2)
library(tidyr)

## Put the dataframes in a list
df_list_goi <- list(GSE4302 = GSE4302_CS_NS_expr_matrix_genes_of_interest, GSE7895 = GSE7895_CS_NS_expr_matrix_genes_of_interest, GSE19027 = GSE19027_CS_NS_expr_matrix_genes_of_interest, GSE20257 = GSE20257_CS_NS_expr_matrix_genes_of_interest, GSE63127 = GSE63127_CS_NS_expr_matrix_genes_of_interest)

## Put the indexes into lists
current_smoker_indexes <- list(GSE4302_csi = GSE4302_current_smoker_indexes, GSE7895_csi = GSE7895_current_smoker_indexes, GSE19027_csi = GSE19027_current_smoker_indexes, GSE20257_csi = GSE20257_current_smoker_indexes, GSE63127_csi = GSE63127_current_smoker_indexes)

never_smoker_indexes <- list(GSE4302_nsi = GSE4302_never_smoker_indexes, GSE7895_nsi = GSE7895_never_smoker_indexes, GSE19027_nsi = GSE19027_never_smoker_indexes, GSE20257_nsi = GSE20257_never_smoker_indexes, GSE63127_nsi = GSE63127_never_smoker_indexes)

# Set current gene of interest (make a loop eventually)

for (i in 1:length(genes_of_interest)){
  
  curr_gene <- genes_of_interest[i] # Loop through indexes of genes of interest
  combined_data_curr_gene <- list(0) # Create a list to hold the data for each dataset for the gene
  
  for (j in 1:length(df_list_goi)){
    
    # Subset data for current smokers and never smokers
    df_curr <- as.data.frame(df_list_goi[j])
    colnames(df_curr) <- sub(".*\\.", "", colnames(df_curr))
    
    csi_curr <- unlist(current_smoker_indexes[j])
    nsi_curr <- unlist(never_smoker_indexes[j])
    
    if(curr_gene %in% df_curr$gene){
      expr_curr_gene <- df_curr[df_curr$gene==curr_gene,]
      expr_curr_gene <- expr_curr_gene[,-1]
      
      expr_current_smokers_curr_gene <- expr_curr_gene[,csi_curr]
      expr_never_smokers_curr_gene <- expr_curr_gene[,nsi_curr]
      
      # Combine into a long format dataframe
      expr_long <- data.frame(
        Gene = curr_gene,
        current_smokers = as.vector(expr_current_smokers_curr_gene),
        never_smokers   = as.vector(expr_never_smokers_curr_gene)
      ) %>%
        pivot_longer(cols = -Gene, names_to = "Group", values_to = "Expression")
      
      expr_long$Group <- sub("\\..*", "", expr_long$Group) # Remove the sample ID info
      
      # Plot using ggplot2
      boxplot<-ggplot(expr_long, aes(x = Group, y = Expression, fill = Group)) +
        geom_boxplot() +
        #facet_wrap(~ Gene, scales = "free_y") +  # One boxplot per gene
        theme_minimal() +
        scale_y_continuous(
          #limits = c(0,14), 
           minor_breaks = NULL) + 
        labs(title = paste0(names(df_list_goi[j]), " ", curr_gene, ": Current vs Never Smokers expression"),
             x = "Group", y = "Expression") +
        theme(legend.position = "none")
      print(boxplot)
    } else{
      print(paste0("Gene ", curr_gene, " not in the dataset ", names(df_list_goi[j])))
    }
  }
}
```



