---
title: "Dataset Intersections"
output: html_document
date: "2024-12-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

2024/12/02: Now unlike the way I did it in the big combined script, here I am going to try out doing more stringent FDR cutoffs as well as fairly relaxed log2FC cutoffs. I may also change the methylation list input, in fact I probably should (either use limma with cbioportal or confirm results from the probe-wise version).

# Comparing and filtering A1, TE, TM, and A2 to generate "linked genes" and "linked persistent genes" lists

## Reading in datasets to intersect
```{r}
GSE63127_CS_NS_GEO2R_limma_sig <- read.table("../2_Outputs/1_Airway_DEGs/GSE63127_CS_NS_GEO2R_limma_sig_20241115.txt", header = TRUE)

DMeth_LUAD_T_NM_hm450_sig <- read.table("../2_Outputs/5_Tumor_DMGs/DMeth_LUAD_T_NM_hm450_sig_20241108_PM.txt")

DGE_LUAD_T_NM_DESeq2_sig <- read.table("../2_Outputs/4_Tumor_DEGs/DGE_LUAD_T_NM_DSeq2_sig_20241127.txt")

TCGA_LUAD_limma_DMG <- read.table("../2_Outputs/5_Tumor_DMGs/TCGA_LUAD_limma_DMG_20241203.txt")

GSE7895_persistent_DEGs <- read.table("../2_Outputs/1_Airway_DEGs/GSE7895_persistent_DEGs_20241127.txt")

```

#######

# I. Using more stringent FDR cutoffs

## I.i Applying FDR cutoffs
```{r}

### Testing FDR and log2FC cutoffs ###

# A1 DEGs
FDR_cutoff_A1=0.001
log2FC_cutoff_A1=0.5
nrow(GSE63127_CS_NS_GEO2R_limma_sig[GSE63127_CS_NS_GEO2R_limma_sig$adj.P.Val<FDR_cutoff_A1
                                    & abs(GSE63127_CS_NS_GEO2R_limma_sig$logFC)>log2FC_cutoff_A1,])

hist(GSE63127_CS_NS_GEO2R_limma_sig[GSE63127_CS_NS_GEO2R_limma_sig$adj.P.Val<FDR_cutoff_A1
                                    & abs(GSE63127_CS_NS_GEO2R_limma_sig$logFC)>log2FC_cutoff_A1,]$logFC, breaks = 100)

# TCGA-LUAD DEGs
FDR_cutoff_TE=0.001
log2FC_cutoff_TE = 1.5

nrow(DGE_LUAD_T_NM_DESeq2_sig[DGE_LUAD_T_NM_DESeq2_sig$FDR<FDR_cutoff_TE
                              & abs(DGE_LUAD_T_NM_DESeq2_sig$log2FC)>log2FC_cutoff_TE,])

hist(DGE_LUAD_T_NM_DESeq2_sig[DGE_LUAD_T_NM_DESeq2_sig$FDR<FDR_cutoff_TE
                              & abs(DGE_LUAD_T_NM_DESeq2_sig$log2FC)>log2FC_cutoff_TE,]$log2FC, breaks = 100)


# TCGA-LUAD DMGs limma

FDR_cutoff_TM=0.001
log2FC_cutoff_TM = 0.5
nrow(TCGA_LUAD_limma_DMG[TCGA_LUAD_limma_DMG$FDR<FDR_cutoff_TM
                               & abs(TCGA_LUAD_limma_DMG$log2FC)>log2FC_cutoff_TM,])

hist(TCGA_LUAD_limma_DMG[TCGA_LUAD_limma_DMG$FDR<FDR_cutoff_TM
                               & abs(TCGA_LUAD_limma_DMG$log2FC)>log2FC_cutoff_TM,]$log2FC, breaks = 100)



### Applying log2FC cutoffs and formatting consistently for merge ###

GSE63127_CS_NS_GEO2R_limma_sig_cutoff <- GSE63127_CS_NS_GEO2R_limma_sig %>%
  filter(abs(logFC)>log2FC_cutoff_A1) %>%
  filter(adj.P.Val < FDR_cutoff_A1) %>%
  dplyr::rename(., Gene = Gene.symbol, FDR_A1 = adj.P.Val, log2FC_A1 = logFC) %>%
  dplyr::select(., Gene, FDR_A1, log2FC_A1)

DGE_LUAD_T_NM_DESeq2_sig_cutoff <- DGE_LUAD_T_NM_DESeq2_sig %>%
  filter(abs(log2FC)>log2FC_cutoff_TE) %>%
  filter(FDR < FDR_cutoff_TE) %>%
  dplyr::rename(., FDR_TE = FDR, log2FC_TE = log2FC, Gene = gene) %>%
  dplyr::select(., Gene, FDR_TE, log2FC_TE)

TCGA_LUAD_limma_DMG_cutoff <- TCGA_LUAD_limma_DMG %>%
  filter(abs(log2FC)>log2FC_cutoff_TM) %>%
  filter(FDR < FDR_cutoff_TM) %>%
  dplyr::rename(., FDR_TM = FDR, log2FC_TM = log2FC) %>%
  dplyr::select(., Gene, FDR_TM, log2FC_TM)



```



### I.ii Intersection of A1 and TE DEGs
```{r}
## Merge the lists
# A1_TE_merged_DEGs <- GSE63127_CS_NS_GEO2R_limma_sig_cutoff %>%
#   inner_join(., DGE_LUAD_T_NM_signed_rank_sig_cutoff, by = "Gene") %>%
#   filter(sign(log2FC_A1)==sign(log2FC_TE)) # Filter to genes with same signs
#   
# nrow(A1_TE_merged_DEGs)

#2024/11/27: Using the DESeq2 list instead
A1_TE_merged_DEGs <- GSE63127_CS_NS_GEO2R_limma_sig_cutoff %>%
  inner_join(., DGE_LUAD_T_NM_DESeq2_sig_cutoff, by = "Gene") %>%
  filter(sign(log2FC_A1)==sign(log2FC_TE)) # Filter to genes with same signs

nrow(A1_TE_merged_DEGs)

```

### I.iii Intersection of A1/TE DEGs with TM DMGs
```{r}
## Merge the lists
# A1_TE_TM_linked_genes <- A1_TE_merged_DEGs %>%
#   inner_join(., DMeth_LUAD_T_NM_hm450_sig_cutoff, by = "Gene") %>%
#   filter(sign(log2FC_A1)!=sign(log2FC_TM)) # Filter to genes with opposite sign
#   
# nrow(A1_TE_TM_linked_genes)

## Using the limma derived DMGs instead
A1_TE_TM_linked_genes <- A1_TE_merged_DEGs %>%
  inner_join(., TCGA_LUAD_limma_DMG_cutoff, by = "Gene") %>%
  filter(sign(log2FC_A1)!=sign(log2FC_TM)) # Filter to genes with opposite sign

nrow(A1_TE_TM_linked_genes)

```


### I.iv.1 Visualizing Spearman correlations of log2FC values
```{r}
## Visualizing Spearman correlations
library(GGally)
ggpairs(A1_TE_TM_linked_genes[, c("log2FC_A1", "log2FC_TE", "log2FC_TM")],
  upper = list(continuous = wrap("cor", method = "spearman")),
  lower = list(continuous = "points", combo = "facethist", discrete = "facetbar", na ="na"),
  diag = list(continuous = "densityDiag", discrete = "barDiag", na = "naDiag"))

```

### I.iv.2 Visualizing multiple linear regression of the log2FC values
```{r}
mlr_airway_model <- lm(log2FC_A1 ~ log2FC_TE + log2FC_TM, data = A1_TE_TM_linked_genes)
summary(mlr_airway_model)


library(scatterplot3d)
s3d <- scatterplot3d(x = A1_TE_TM_linked_genes$log2FC_A1, 
              y = A1_TE_TM_linked_genes$log2FC_TE, 
              z = A1_TE_TM_linked_genes$log2FC_TM,
              main="Plotting airway expression vs. tumor expression vs. tumor methylation",
              xlab = "log2FC(TE)",
              ylab = "log2FC(TM)",
              zlab = "log2FC(A1)",
              pch = 19,
              #color = colors_3d, # getting color values from methylation sign
              color = "steelblue",
              labels = A1_TE_TM_linked_genes$Gene,
              type = "h",
              #highlight.3d = TRUE, 
              angle = 60
              )
s3d$plane3d(mlr_airway_model) # Add a plane based on the multiple linear regression model

```

OK here is an amazing interaction network I got:

https://version-12-0.string-db.org/cgi/network?networkId=bK3IJNEMHJEy

This was using the 29 genes given by the parameters:

FDR_cutoff_A1=0.001
log2FC_cutoff_A1=0.5

FDR_cutoff_TE=0.001
log2FC_cutoff_TE = 1.5

FDR_cutoff_TM=0.001
log2FC_cutoff_TM = 0.5

Now to check via intersection with the "persistent" genes:


## 5.3 Comparing and filtering A2 and "linked" genes to give "persistent linked genes"

### 5.3.1 Applying log2FC cutoffs
```{r}
# For now, no cutoffs
```

### 5.3.2 Filtering by dataset intersection
```{r}

A1_TE_TM_A2_persistent_linked_genes <- A1_TE_TM_linked_genes %>%
  dplyr::inner_join(., GSE7895_persistent_DEGs, by = "Gene") %>%
  dplyr::filter(sign(log2FC_A1) == sign(CS_NS_A2)) # Ensure signs are the same between A1 and A2 CS vs NS

nrow(A1_TE_TM_A2_persistent_linked_genes)

A1_TE_TM_A2_persistent_linked_genes

## Recording results based on cutoffs (when I was using the signed-rank test for TE - now I use DESeq2)
## 0.5,1,0.5,0 => 1 (NQO1)
## 0,0,0,0 => 27
## 0.2, 1, 0.3 => 10

```

### 5.3.3 Additional checks (correlation of log2FCs in A1 and A2)
```{r}
cor.test(A1_TE_TM_A2_persistent_linked_genes$log2FC_A1, y = A1_TE_TM_A2_persistent_linked_genes$CS_NS_A2, use = "everything",
    method = "spearman")
    
ggplot(A1_TE_TM_A2_persistent_linked_genes, aes(x = log2FC_A1, y = CS_NS_A2)) +
  geom_point(color = "blue", size = 3) +                # Scatter points
  geom_smooth(method = "lm", se = FALSE, color = "black") + # Add a trend line +
  theme_minimal()
```
