---
title: "5_Gene_list_comparisons"
output: html_document
date: "2024-10-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Comparing the output lists of "linked genes" and "linked persistent genes" to gene lists from other types of unique studies that relate to the question of persistent/irreversible effects in airways.

## Reading in the current versions of the "linked genes" and "linked persistent genes" (subject to change)
```{r}

# linked_DEGs_v7 <- read.table("../1_Comparison_output/Aim1_Aim2_linked_DEGs_v7.txt", sep = '\t', header = TRUE)
# 
# persistent_linked_DEGs_v7 <- read.table("../1_Comparison_output/14_genes_pipeline_v7.txt", sep = '\t')
# colnames(persistent_linked_DEGs_v7) <- "Gene_Symbol"

linked_DEGs <- read.table("../2_Outputs/A1_TE_TM_linked_genes_nocutoffs_20241112.txt", sep = '\t', header = TRUE)

persistent_linked_DEGs <- read.table("../2_Outputs/A1_TE_TM_A2_persistent_linked_genes_nocutoffs_20241112.txt", sep = '\t', header = TRUE)

```

## Compare linked gene lists with studies related to irreversible expression changes in airways
```{r}

## Chari et al 2007 (Lam lab SAGE study) ##

raj_2007_irreversible_gene_list <- read.table("../3_Gene_lists/irreversible_expression_airways/17727719_raj_2007/17727719_raj_2007_irreversible_gene_list.txt")
colnames(raj_2007_irreversible_gene_list) <- "Gene_Symbol"

print(c("Genes common to Chari 2007 irreversible gene list and the linked DEGs: ", intersect(linked_DEGs$Gene, raj_2007_irreversible_gene_list$Gene)))

print(c("Genes common to Chari 2007 irreversible gene list and the persistent linked DEGs: ", intersect(persistent_linked_DEGs$Gene, raj_2007_irreversible_gene_list$Gene)))

## Barel et all 2023 (Cornell lab serial sampling study)

barel_2023_irreversible_gene_list <- read.table("../3_Gene_lists/irreversible_expression_airways/37531632_barel_2023 /Table_2_truncated_37531632.txt", header = TRUE)

print(c("Genes common to Barel 2023 serial sampling irreversible gene list and the linked DEGs: ", intersect(linked_DEGs$Gene, barel_2023_irreversible_gene_list$Gene)))

print(c("Genes common to Barel 2023 serial sampling irreversible gene list and the persistent linked DEGs: ", intersect(persistent_linked_DEGs$Gene, barel_2023_irreversible_gene_list$Gene)))

## (For sake of completeness do the reference study later. Also make venn diagrams of this.)

```

## Compare linked gene lists with studies related to methylation in airways

```{r}

## Vucic et al 2014 (link to COPD) ##

# Read in table 2 ("Differentially Methylated and Differentially Expressed Genes in Chronic Obstructive Pulmonary Disease Small Airways Previously Associated with Chronic Obstructive Pulmonary Disease")
vucic_2014_table_2 <- read.table("../3_Gene_lists/methylation_airways/24298892/24298892_vucic_2014_table_2", sep = '\t', header = TRUE)

print(c("Genes common to Vucic 2014 known COPD DMG list and the linked DEGs: ", intersect(linked_DEGs$Gene, vucic_2014_table_2$Symbol)))

print(c("Genes common to Vucic 2014 known COPD DMG list and the persistent linked DEGs: ", intersect(persistent_linked_DEGs$Gene, vucic_2014_table_2$Symbol)))

# Read in table 3 ("Differentially Methylated and Expressed Genes Most Likely Under Epigenetic Control in Chronic Obstructive Pulmonary Disease Small Airways")
vucic_2014_table_3 <- read.table("../3_Gene_lists/methylation_airways/24298892/24298892_vucic_2014_table_3", sep = '\t', header = TRUE)

print(c("Genes common to Vucic 2014 COPD-related linked DEG-DMG list and the linked DEGs: ", intersect(linked_DEGs$Gene, vucic_2014_table_3$Symbol)))

print(c("Genes common to Vucic 2014 COPD-related linked DEG-DMG  list and the persistent linked DEGs: ", intersect(persistent_linked_DEGs$Gene, vucic_2014_table_3$Symbol)))



## Buro-Auriemma et al 2013 (smoking-related airway methylation changes)##

Buro_2013_25_hypermethylated <- read.table("../3_Gene_lists/methylation_airways/23842454/23842454_Buro_2013_25_hypermethylated.txt")

print(c("Genes common to Buro-Auriemma smoking-related hypermethylated gene list and the linked DEGs:", intersect(linked_DEGs$Gene, Buro_2013_25_hypermethylated$V1)))

print(c("Genes common to Buro-Auriemma smoking-related hypermethylated gene list and the persistent linked DEGs: ", intersect(persistent_linked_DEGs$Gene, Buro_2013_25_hypermethylated$V1)))

Buro_2013_25_hypomethylated <- read.table("../3_Gene_lists/methylation_airways/23842454/23842454_Buro_2013_25_hypomethylated.txt")

print(c("Genes common to Buro-Auriemma 2013 smoking-related hypomethylated gene list and the linked DEGs:", intersect(linked_DEGs$Gene, Buro_2013_25_hypomethylated$V1)))

print(c("Genes common to Buro-Auriemma 2013 smoking-related hypomethylated gene list and the persistent linked DEGs: ", intersect(persistent_linked_DEGs$Gene, Buro_2013_25_hypomethylated$V1)))


```

## Compare with study of irreversible expression changes in tumour-adjacent normal tissue
```{r}

# Read in Table 3 from Bosse et al 2012 ("Replicated probe sets upregulated by smoking and not returning to never-smoker levels within 10 years of smoking cessation in the discovery set and validated in at least one of the 2 replication sets")

bosse_2012_table_3 <- read.table("../3_Gene_lists/tumor_adjacent_normal/22659451_bosse_2012_table_3.txt", header = TRUE)

print(c("Genes common to Bosse 2012 list of genes persistently upregulated in tumor-adjacent normal tissue 10 years after smoking cessation, and the linked DEGs list:", intersect(linked_DEGs$Gene, bosse_2012_table_3$Gene_symbol)))

print(c("Genes common to Bosse 2012 list of genes persistently upregulated in tumor-adjacent normal tissue 10 years after smoking cessation, and the persistent linked DEGs list:", intersect(persistent_linked_DEGs$Gene, bosse_2012_table_3$Gene_symbol)))

```

2024/10/09 -- Current version is not looking too interesting unfortunately, except for the few agreements in the irreversibility studies...
Could try expanding the gene list...or checking the miRNA targets somehow...

Airway methylation seems like a big thing, maybe I should try a comparison with the airway expression first as a separate branch?





